Estimating Information Transfer in Hard Parameter Sharing via Precise Bias-variance Asymptotics

### 2021-07-14 revision_journal/setup.tex
finish sec 3.2

\subsection{Data model}

@@

Having defined the data model for each task, next, we describe typical scenarios in transfer learning under our setup.
We define two settings as follows:
\begin{itemize}
    \item \textbf{Covariate shift:} independent $X^{(1)}$ and $X^{(2)}$ with the same population covariance matrices but different models.
    \item \textbf{Model shift:} independent $X^{(1)}$ and $X^{(2)}$ with the same population covariance matrices (i.e. $\Sigma^{(1)}=\Sigma^{(2)}$) but different models (i.e. $\beta^{(1)} \neq \beta^{(2)}$).
\end{itemize}
One interesting special case of capturing different models is the following random-effect model, which has been studied in several recent works concerning high-dimensional (single-task) ridge regression \cite{dobriban2018high} and distributed regression \cite{dobriban2020wonder}.
 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
%\begin{example}%[Sample sizes]
%\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, let
\begin{align}
    \beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, \label{eq_re}
\end{align}
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$.

@@

%We describe our estimators that combine the information from both datasets.
We learn a two-layer linear neural network with parameters $A \in \real^{p\times r}$, $B_1 \in \real^{r}$, and $B_2 \in \real^r$ by minimizing the following optimization objective:
\begin{align}\label{eq_hps}
    f(A, B_1, B_2) = \bignorm{X^{(1)} A B_1 - Y^{(1)}}_2^2 + \bignorm{X^{(2)} A B_2 - Y^{(2)}}_2^2.
\end{align}
The above two-layer neural network (also called hard parameter sharing \cite{C97}) captures the intuition that both datasets share the feature space (i.e. the subspace spanned by the columns of $A$) while each having a separate prediction head.
Suppose that $(\hat{A}, \hat{B_1}, \hat{B_2})$ is a minimizer of $f(\cdot)$.
We define the hard parameter sharing estimator for task $i$ as
\[ \hat{\beta}_i^{\MTL} \define \hat{B} \hat{A}_i. \]
Our study of such estimators is motivated by the fact that hard paratemer sharing is widely used in empirical studies of multi-task learning \cite{R17}.
\todo{Later on in Section \ref{sec_diff}, we will further compare the HPS estimator to other natural transfer learning estimators.} %$L(\hat{\beta}_2^{\STL})$ and $L(\hat{\beta}^{\AV}(\lambda))$ for some properly chosen $\lambda\in [0,1]$.}

We measure the performance of $\hat{\beta}_i^{\MTL}$ through the out-of-sample predication risk (or test loss).
Consider a datapoint $(x,y)$ generated from the same model as dataset $i$: $y= x^\top \beta^{(i)} + \ve$. %, where $x \in \R^p$ and $\ve\in \R$ are independent of $X^{(1)}$, $X^{(2)}$, $\varepsilon^{(1)}$ and $\varepsilon^{(2)}$, and only $x$ is observable.
%We want to use $x^\top \hat{\beta}_2^{\MTL} (\hat a) $ to predict $y$, and we measure
The expected predication risk (under mean squared error) is given by
$$\exarg{x, \varepsilon}{\left\|y-x^\top \hat{\beta}_i^{\MTL} \right\|^2}= \left\|\big(\Sigma^{(i)}\big)^{1/2} \left(\hat{\beta}_i^{\MTL} - \beta^{(i)}\right)\right\|^2 + \sigma^2.$$
%Since $\sigma^2$ is a constant that does not depend on the model, we ignore it and define the predication loss as
The excess risk is defined without the constant term $\sigma^2$:
\be\label{HPS_loss}
    L(\hat{\beta}_i^{\MTL}):= \left\| \big(\Sigma^{(i)}\big)^{1/2} \left(\hat{\beta}_i^{\MTL} - \beta^{(i)}\right)\right\|^2 .
\ee

@@

\begin{proposition}[Optimum under the random-effect model]\label{lem_hat_v}
%Suppose the assumptions of Lemma \ref{prop_model_shift_tight} hold. Assume that $ \kappa^2 \sim pd^2 \sim \sigma^2$ are of the same order.
    Suppose Assumption \ref{assm_big1} and the above random-effect model under equation \eqref{eq_re} holds. %in Example \ref{ex_same_cov} hold.
    Suppose that for a constant $c_0>0$,
	\be\label{para_rel}
	    \|\beta_0\|^2 \ge p^{c_0}d^2 +p^{-1/2+c_0}\sigma^2.
	\ee
    Then we have that for any small constant $c>0$ and large constant $ C>0$,
	%Let $c$ be a sufficiently small fixed constant.
	%In the setting of Corollary \ref{cor_MTL_loss},
%There exists a constant $C>0$ such that
	\be\label{hatw_add1}
	 \hat a =1+ \OO\left(\frac{d^2}{\|\beta_0\|^2} + p^{-1/4+c} \frac{d+\sigma}{\|\beta_0\|} + p^{-C}\right) \quad \text{w.h.p.}
	\ee
\end{proposition}
In this paper, we say that an event $\Xi$ holds \emph{with high probability} (w.h.p.) if $\mathbb P(\Xi)\to 1$ as $p\to \infty$.

%Suppose every $\beta^{(i)}$ consists of two random components, one that is shared among all tasks and one that is task-specific.
%%Thus, each task contributes a certain amount to the shared component and injects a task-specific bias.
%More precisely, %we have
%$$\beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2,\cdots, t,$$
%where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
%and $\wt \beta^{(i)}$ denotes the $i$-th
%%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
%task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$.
%We emphasize that unlike Theorem \ref{cor_MTL_loss}, for Proposition \ref{lem_hat_v} we do not assume $\Sigma^{(1)}=\Sigma^{(2)}$ and the Gaussian distributions of the entries of $Z^{(1)}$ and $Z^{(2)}$.
%In particular, its exact dependence on the model parameters, including the model bias, noise variance, sample sizes, and population covariance matrices, is not well-understood so far. %\FY{What is known so far about this topic?}

@@

%$L(\hat{\beta}_2^{\STL})$ of the single-task OLS estimator  as a baseline.

@@

%The bias equation \eqref{Lbias} captures the 
%Plugging \eqref{HPS_est} into \eqref{HPS_loss}, we get
%\begin{align}
%    L(\hat{\beta}_2^{\MTL}(\hat a))  =  &\left\| (\Sigma^{(2)})^{1/2}\hat \Sigma(\hat a)^{-1} (X^{(1)})^\top X^{(1)} (\hat a\beta^{(1)}-\hat a^2\beta^{(2)}) \right. \nonumber\\
%    &\left. +(\Sigma^{(2)})^{1/2}\hat \Sigma(\hat a)^{-1} \left[ (X^{(2)})^\top \epsilon^{(2)} + \hat a  (X^{(1)})^\top \epsilon^{(1)} \right]\right\|^2.\label{Lbeta_HPS}
%\end{align}
%Note that 
%    is called the variance term, which depends on the noise variance. 
    %predication loss for the HPS and OLS estimators.
%\end{proof}

@@

%In this section, we show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition, when all tasks have the same features.

%\paragraph{Random matrix theory.}
Building on the above connection, we apply recent developments from random matrix theory to analyze information transfer.
As a warmup, we describe the following classical result in multivariate statistics that characterizes the excess risk (or equivalently, the variance) of the OLS estimator.

\begin{lemma}[Variance asymtotics of OLS estimators, cf. Theorem 2.4 of \citet{isotropic} and Theorem 3.14 of \citet{DY}]\label{fact_tr}
	%Let $X  \in \real^{n\times p}$ be a random matrix that satisfies Assumption \ref{assume_rm}.
	%Let $\Sigma\in\real^{p\times p}$ denote the population covariance matrix of $X$.
	Under Assumption \ref{assm_big1}, we have that for any deterministic $p\times p$ matrix $\Sigma$,
		\[ \bigtr{\Sigma \frac1{(X^{(2)})^\top X^{(2)} }  } %= \bigtr{ \frac1{(Z^{(2)})^\top Z^{(2)} }  }
		= \frac{1}{n_2 - p}\bigtr{\Sigma (\Sigma^{(2)})^{-1}} +  \OO\left( \frac{p^{\e}}{\sqrt{n_2 p}}\cdot \frac{p}{n_2}\|\Sigma\|_{op} \right) \]
		with high probability for any small constant $\e>0$.
\end{lemma}
\todo{the scaling $\sqrt{n_2}$ above doesn't seem correct}

>>>>

\subsection{Data model}\label{sec_data}

@@

Having defined the data model for each task, next, we describe typical scenarios in transfer learning under our setup.
We define two settings as follows:
\begin{itemize}
    \item \textbf{Covariate shift:} independent $X^{(1)}$ and $X^{(2)}$ with the same population covariance matrices but different models.
    \item \textbf{Model shift:} independent $X^{(1)}$ and $X^{(2)}$ with the same population covariance matrices (i.e. $\Sigma^{(1)}=\Sigma^{(2)}$) but different models (i.e. $\beta^{(1)} \neq \beta^{(2)}$).
\end{itemize}
One interesting special case of capturing different models is the following random-effect model, which has been studied in several recent works concerning high-dimensional (single-task) ridge regression \cite{dobriban2018high} and distributed regression \cite{dobriban2020wonder}.
 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
%\begin{example}%[Sample sizes]
%\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, let
\begin{align}
    \beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, \label{eq_re}
\end{align}
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} \mu^2$.

@@

%We describe our estimators that combine the information from both datasets.
We learn a two-layer linear neural network with parameters $A \in \real^{p\times r}$, $B_1 \in \real^{r}$, and $B_2 \in \real^r$ by minimizing the following optimization objective:
\begin{align}\label{eq_hps}
    f(A, B_1, B_2) = \bignorm{X^{(1)} A B_1 - Y^{(1)}}_2^2 + \bignorm{X^{(2)} A B_2 - Y^{(2)}}_2^2.
\end{align}
The above two-layer neural network (also called hard parameter sharing \cite{C97}) captures the intuition that both datasets share the feature space (i.e. the subspace spanned by the columns of $A$) while each having a separate prediction head.
Suppose that $(\hat{A}, \hat{B_1}, \hat{B_2})$ is a minimizer of $f(\cdot)$.
We define the hard parameter sharing estimator for task $i$ as
\[ \hat{\beta}_i^{\MTL} \define \hat{B} \hat{A}_i. \]
Our study of such estimators is motivated by the fact that hard paratemer sharing is widely used in empirical studies of multi-task learning \cite{R17}.
\todo{Later on in Section \ref{sec_diff}, we will further compare the HPS estimator to other natural transfer learning estimators.}

We measure the performance of $\hat{\beta}_i^{\MTL}$ through the out-of-sample predication risk (or test loss).
Consider a datapoint $(x,y)$ generated from the same model as dataset $i$: $y= x^\top \beta^{(i)} + \ve$.
The expected predication risk (under mean squared error) is given by
$$\exarg{x, \varepsilon}{\left\|y-x^\top \hat{\beta}_i^{\MTL} \right\|^2}= \left\|\big(\Sigma^{(i)}\big)^{1/2} \left(\hat{\beta}_i^{\MTL} - \beta^{(i)}\right)\right\|^2 + \sigma^2.$$
The excess risk is defined without the constant term $\sigma^2$:
\be\label{HPS_loss}
    L(\hat{\beta}_i^{\MTL}):= \left\| \big(\Sigma^{(i)}\big)^{1/2} \left(\hat{\beta}_i^{\MTL} - \beta^{(i)}\right)\right\|^2 .
\ee

@@

\begin{proposition}[Optimum under the random-effect model]\label{lem_hat_v}
%Suppose the assumptions of Lemma \ref{prop_model_shift_tight} hold. Assume that $ \kappa^2 \sim pd^2 \sim \sigma^2$ are of the same order.
    Suppose Assumption \ref{assm_big1} and the above random-effect model under equation \eqref{eq_re} holds. %in Example \ref{ex_same_cov} hold.
    Suppose that for a constant $c_0>0$,
	\be\label{para_rel}
	    \|\beta_0\|^2 \ge p^{c_0}\mu^2 +p^{-1/2+c_0}\sigma^2.
	\ee
    Then we have that for any small constant $c>0$ and large constant $ C>0$,
	%Let $c$ be a sufficiently small fixed constant.
	%In the setting of Corollary \ref{cor_MTL_loss},
%There exists a constant $C>0$ such that
	\be\label{hatw_add1}
	 \hat a =1+ \OO\left(\frac{\mu^2}{\|\beta_0\|^2} + p^{-\frac 1 4 + c} \frac{\mu+\sigma}{\|\beta_0\|} + p^{-C}\right) \quad \text{w.h.p.}
	\ee
\end{proposition}
In this paper, we say that an event $\Xi$ holds \emph{with high probability} (w.h.p.) if $\mathbb P(\Xi)\to 1$ as $p\to \infty$.
\todo{change $d$ to $\mu$ because a reviewer complained}
%Suppose every $\beta^{(i)}$ consists of two random components, one that is shared among all tasks and one that is task-specific.
%%Thus, each task contributes a certain amount to the shared component and injects a task-specific bias.
%More precisely, %we have
%$$\beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2,\cdots, t,$$
%where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
%and $\wt \beta^{(i)}$ denotes the $i$-th
%%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
%task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$.
%We emphasize that unlike Theorem \ref{cor_MTL_loss}, for Proposition \ref{lem_hat_v} we do not assume $\Sigma^{(1)}=\Sigma^{(2)}$ and the Gaussian distributions of the entries of $Z^{(1)}$ and $Z^{(2)}$.
%In particular, its exact dependence on the model parameters, including the model bias, noise variance, sample sizes, and population covariance matrices, is not well-understood so far. %\FY{What is known so far about this topic?}

@@

@@

%The bias equation \eqref{Lbias} captures the 
%Plugging \eqref{HPS_est} into \eqref{HPS_loss}, we get
%\begin{align}
%    L(\hat{\beta}_2^{\MTL}(\hat a))  =  &\left\| (\Sigma^{(2)})^{1/2}\hat \Sigma(\hat a)^{-1} (X^{(1)})^\top X^{(1)} (\hat a\beta^{(1)}-\hat a^2\beta^{(2)}) \right. \nonumber\\
%    &\left. +(\Sigma^{(2)})^{1/2}\hat \Sigma(\hat a)^{-1} \left[ (X^{(2)})^\top \epsilon^{(2)} + \hat a  (X^{(1)})^\top \epsilon^{(1)} \right]\right\|^2.\label{Lbeta_HPS}
%\end{align}

@@

Building on the above connection, we apply recent developments from random matrix theory to analyze information transfer.
As a warmup, we describe the following classical result in multivariate statistics that characterizes the excess risk (or equivalently, the variance) of the OLS estimator.

\begin{lemma}[Variance asymtotics of OLS estimators, cf. Theorem 2.4 of \citet{isotropic} and Theorem 3.14 of \citet{DY}]\label{fact_tr}
	%Let $X  \in \real^{n\times p}$ be a random matrix that satisfies Assumption \ref{assume_rm}.
	%Let $\Sigma\in\real^{p\times p}$ denote the population covariance matrix of $X$.
	Under Assumption \ref{assm_big1}, we have that for any deterministic  matrix $\Sigma\in\real^{p\times p}$,
		\[ \bigtr{\Sigma \Big((X^{(2)})^\top X^{(2)} \Big)^{-1}  } %= \bigtr{ \frac1{(Z^{(2)})^\top Z^{(2)} }  }
		= \frac{1}{n_2 - p}\bigtr{\Sigma (\Sigma^{(2)})^{-1}} +  \OO\left( \frac{p^{\e}}{\sqrt{n_2 p}}\cdot \frac{p}{n_2}\|\Sigma\|_{op} \right) \]
		with high probability for any small constant $\e>0$.
\end{lemma}
\todo{the scaling $\sqrt{n_2}$ above doesn't seem correct}
